Against Spurious Understanding: Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}

### 2024-02-06 appendix.tex
removed spaces from file names for arxiv

%\subsection{Example: Autoencoder}\label{example-deep-learning}
\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}

%So far we have considered simple linear transformations of data. One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning. Our next example will therefore involve a deep neural network.

This is an additional example that we have not discuss in the body of the paper. Here, we build forth on an application in Economics. However, we now seek to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks. For this, we will use a neural network architecture.
%, that allows us to recover a compact latent representation of the yield curve.

@@

%Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\). Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\). Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).

%To do this, we use a simple autoencoder architecture. With this
Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\). Finally, we let \(\theta\) denote our model parameters. Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).

%To do this, we use a simple autoencoder architecture: 
The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function. The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons. The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data. The output layer consists of a single neuron for our output variable, \(g_t\). We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.

@@

Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve? To us, that would be too big of a statement. Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
%It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data. But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve. While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
%\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model. To demonstrate this, let us consider a simple linear regression model for GDP growth. We might be interested in understanding to what degree economic growth in the past is associated with economic growth today. As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient. However, this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.

% A less risky/strong term than confounding bias that we could use here is omitted variable bias. Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI (n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_PPI (n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Mo) (n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Yr) (n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (10 Yr) (n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI (n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI (n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Mo).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Yr).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (10 Yr).png}

>>>>

\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}

This is an additional example that we have not discuss in the body of the paper. Here, we build forth on an application in Economics. However, we now seek to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks. For this, we will use a neural network architecture.

@@

Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\). Finally, we let \(\theta\) denote our model parameters. Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).

The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function. The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons. The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data. The output layer consists of a single neuron for our output variable, \(g_t\). We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.

@@

Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve? To us, that would be too big of a statement. Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models. For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model. To demonstrate this, let us consider a simple linear regression model for GDP growth. We might be interested in understanding to what degree economic growth in the past is associated with economic growth today. As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient. However, this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI_(n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_PPI_(n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_(1Mo)_(n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_(1Yr)_(n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_(10Yr)_(n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI_(n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI_(n_pc=128).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_(1Mo).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_(1Yr).png}

@@

\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_(10Yr).png}
